Interpolation of Sparse High-Dimensional Data\thanks {This work was supported by the National Science Foundation Grant CNS-1565314.}

### 2020-08-13 [2019-02]_NUMA_(interpolant_error_bound)/tchlux-NA-interpolation/Versions/3-paper.tex
Finished NUMA paper and submitted. Might need to do one more round with editors.

%% Include the response to reviewers at the beginning of the document.
\include{tchlux-NA-response-2020-04}

@@

Regression and interpolation have a considerable theoretical base in
one dimension \cite{cheney2009course}. \changed{Common techniques
  include Lagrangian interpolation, piecewise linear interpolation,
  and more generally spline interpolation via B-splines
  \cite{de1978practical}}{Splines in particular are well understood
  as an interpolation technique in one dimension
  \cite{de1978practical}, particularly B-splines}. Tensor products of
B-splines \cite{unther1996interpolating} or other basis functions have
an unfortunate exponential scaling with increasing
dimension. Exponential scaling prohibits tensor products from being
reasonably applied beyond three-dimensional data. In order to address
this dimensional scaling challenge, C. de Boor and others proposed box
splines \cite{de2013box}, of which one of the approximation techniques
in this work is composed \cite{lux2018novel}.

The theoretical foundation of low dimensional interpolation allows the
construction of strong error bounds that are \changed{lacking
  in}{absent from} high dimensional problems. \changed{A large body of
  literature exists studying the construction and application of
  sparse grids \cite{bungartz2004sparse,chkifa2014high}, which attempt
  to address the problem of dimensional scaling. The sparse grid point
  sets combined with specific approximation functions are used to
  construct total error bounds by considering projections onto
  orthogonal polynomials
  \cite{migliorati2013approximation,nobile2016convergence}. However,
  sparse grids have exponential size in dimension making them
  intractable for more than tens of dimensions and do not immediately
  apply to the problem of scattered data interpolation.}{} This work
extends some known results regarding the secant method
\cite{dennis1996numerical} to construct \changed{a uniform error bound
  for interpolants over arbitrary scattered data in any dimension}{an
  interpolation error bound for problems of arbitrary
  dimension}. These error bounds are useful, considering
\changed{similar uniform bounds}{the same cannot} cannot be
\changed{constructed}{said} for regression algorithms in general. The
maximum complexity of an interpolant is bounded by the amount of data
available, while the maximum complexity of a regressor is bounded by
both the amount of data and the chosen parametric form. For this
reason, generic uniform bounds are largely unobtainable for regression
techniques on arbitrary approximation problems, even when the
approximation domain is bounded.

@@

The most challenging problem when scaling in dimension is that the
number of possible interactions between dimensions grows
exponentially. Quantifying all possible interactions becomes
intractable, and hence \changed{problems}{} beyond \changed{ten to
  twenty dimensions usually rely on linear models}{three-dimensional
  data mostly linear models are used}. That is not to say nonlinear
models are absent, but nonlinearities are often either preconceived or
model \changed{interactions between small subsets of}{pairwise
  interactions between} dimensions at most. Even globally nonlinear
approximations such as neural networks are constructed from
compositions of summed low-interaction functions
\cite{clevert2015fast}.

@@

\subsection{Multivariate Adaptive Regression Splines}
\label{sec:mars}
This approximation was introduced in \cite{friedman1991multivariate}
and subsequently improved to its current version in
\cite{stanford1993fast}, called fast multivariate adaptive regression
splines (Fast MARS). In Fast MARS, a least squares fit model is
iteratively built by beginning with a single constant valued function
and adding two new basis functions at each iteration of the form
\begin{align*}
  B_{2j-1}(x) &= B_l(x) \bigl(x_i-x^{(p)}_i\bigr)_+, \\
  B_{2j}(x) &= B_k(x) \bigl(x_i-x^{(p)}_i\bigr)_- ,
\end{align*}
where \changed{$j \leq m$ is the iteration number, $m$ is the maximum
  number of underlying basis functions}{}, $1 \le p \le n$, and
$B_l(x)$, $B_k(x)$ are basis functions from the previous iteration,
 $$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$ and
$w_- = (-w)_+$. After iteratively constructing a model, MARS then
iteratively removes basis functions that do not contribute to goodness
of fit. In effect, MARS creates a locally component-wise tensor
product approximation of the data. The overall computational
complexity of Fast MARS is $\mathcal{O}(n d m^3)$ \changed{}{where $m$
  is the maximum number of underlying basis functions}. This paper
uses an implementation of Fast MARS \cite{rudy2017pyearth} with $m =
200$ throughout all experiments.

@@

\subsection{Multilayer Perceptron Regressor}
\label{sec:mlp}
The neural network is a well studied and widely used method for both
regression and classification tasks
\cite{hornik1989multilayer,rumelhart1988learning}. When using the
rectified linear unit (ReLU) activation function
\cite{dahl2013improving} and fitting the model with stochastic
gradient descent (SGD) or BFGS minimization techniques
\cite{goh2017why,moller1993scaled,robbins1951stochastic}, the model
built by a multilayer perceptron uses layers $l : \mathbb{R}^{i}
\rightarrow \mathbb{R}^{j}$ defined by
 $$ l(u) = \bigl( u^t W_l + c_l \bigr)_+ ,$$
where $c_l \in \mathbb{R}^j$ and $W_l$ is the $i \times j$ weight
matrix for layer $l$. In this form, the multilayer perceptron (MLP)
produces a piecewise linear model of the input data. The computational
complexity of training a multilayer perceptron is $\mathcal{O}(n d
m)$, where $m$ is determined by the sizes of the layers of the network
and the stopping criterion of the minimization used for finding
weights. This paper uses an MLP built from Keras and Tensorflow to
perform regression \cite{tensorflow2015-whitepaper,chollet2015keras}
with ten hidden layers each having thirty two nodes (a total of
approximately ten thousand parameters), ReLU activation, and
\changed{one hundred}{five} thousand \changed{steps}{epochs} of SGD
for training. \changed{It should be noted that the choice of neural
  network architecture impacts approximation performance, but no
  architecture tuning is performed here}{}.

@@

The Delaunay triangulation is a well-studied geometric technique for
producing an interpolant \cite{lee1980two}. The Delaunay triangulation
of a set of data points into simplices is such that there are no data
points inside the sphere defined by the vertices of each simplex. For
a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$,
$x^{(d)}$, and function values $f\bigl(x^{(i)}\bigr)$, $i=0$,
$\ldots$, $d$, $y \in S$ is a unique convex combination of the
vertices,
 $$ y = \sum_{i=0}^{d} w_i x^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad w_i \geq 0, \quad i=0,\ldots,d, $$
and the Delaunay interpolant $\hat f(y)$ at $y$ is given by
 $$ \hat f(y) = \sum_{i=0}^{d} w_i f\bigl(x^{(i)}\bigr). $$ The
computational complexity of Delaunay interpolation (for the
implementation used \cite{chang2018polynomial}) is $\mathcal{O}(k n
d^3)$. Given pathological data the entire triangulation could be
computed with $k = n^{\lceil d / 2 \rceil}$, but the average case yields
$k = d \log d$ and is capable of scaling reasonably to $d \leq 50$. In
the present application, a Delaunay simplex $S$ containing $y$ is
found, then the $d+1$ vertices (points in $X$) of $S$ are used to
assign weights to each vertex and produce the predicted function value
for point $y$.

\changed{
\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth,]{example_delaunay.pdf}
  \includegraphics[width=0.4\textwidth,]{example_not_delaunay.pdf}
  \caption{\color{red} On the left above is a depiction of a Delaunay
    triangulation over four points, notice that the circumball (shaded
    circle) for the left simplex does not contain the fourth point.
    On the right above, a non-Delaunay mesh is depicted. Notice that
    the circumball for the top simplex (shaded circle, clipped at
    bottom edge of the visual) contains the fourth point which
    violates the Delaunay condition for a simplex.
  \vspace{-.1cm}}
  \label{fig:delaunay}
\end{figure}
}{}

@@

The modified Shepard method used here (ShepMod) generates a continuous
approximation based on Euclidean distance and resembles a nearest
neighbor interpolant \cite{cover1967nearest}. This model is a type of
\textit{metric interpolation}, also called a Shepard method
\cite{gordon1978shepard,shepard1968two}. The interpolant has the form
 $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}
     {\sum\limits_{k=1}^{n}W_k(x)} ,$$
where \changed{$W_k(x^{(k)}) = 1$, $W_k(x) = \bigl(\bigl(r_k -
  \bigl\|x - x^{(k)}\bigr\|_2\bigr)_+ \big/ \bigl( r_k \bigl\|x -
  x^{(k)}\bigr\|_2 \bigr)\bigr)^2$ for $x \neq x^{(k)}$,}{} and $r_k
\in \mathbb{R}$ is the smallest radius such that at least $d+1$ other
points $x^{(j)}$, $j \not = k$, are inside the closed Euclidean ball
of radius $r_k$ about $x^{(k)}$. The computational complexity of
ShepMod is $\mathcal{O}(n^2d)$. This paper uses a Fortran 95
implementation of ShepMod derived from SHEPPACK
\cite{thacker2010algorithm}.

\changed{
\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth,]{example_shepard_with_weight.pdf}
  \caption{\color{red} On the left above is a depiction of the radius
    of influence for three chosen points of a collection in two
    dimensions using the modified Shepard criteria. On the right is
    the shape of the weight function $W_k$ for all data points, where
    $W_k(x) = 0$ when $\|x-x^{(k)}\|_2 \geq r_k$. It can also be seen
    that in the limit as $W_k$ tends towards infinity at $x_k$, the
    relative weight of $x_k$ tends to one.
  \vspace{-.1cm}}
  \label{fig:shepard}
\end{figure}
}{}

\subsection{Linear Shepard}
The linear Shepard method (LSHEP) is a blending function using local
linear interpolants, a special case of the general Shepard algorithm
\cite{thacker2010algorithm}. The interpolant has the form
 $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)P_k(x)}
 {\sum\limits_{k=1}^{n}W_k(x)} ,$$
where $W_k(x)$ is \changed{the same as for the modified Shepard
  method}{a locally supported weighting function} and $P_k(x)$ is a
local linear approximation to the data satisfying
$P_k\bigl(x^{(k)}\bigr) = f\bigl(x^{(x)}\bigr)$. The computational
complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This paper uses the
Fortran 95 implementation of LSHEP in SHEPPACK
\cite{thacker2010algorithm}.

@@

\changed{
\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth,]{VM_Construction.pdf}
  \caption{\color{red} Above is a depiction of the Voronoi cell
    boundaries (dashed lines) about a set of interpolation points
    (dots) in two dimensions. In this example, the Voronoi model basis
    function about the center most point has nonzero weight in the
    shaded region and transitions from a value of one at the point to
    zero at the boundary of the twice expanded Voronoi cell (solid
    line).
  \vspace{-.1cm}}
  \label{fig:voronoi-basis}
\end{figure}
}{}

The nearest neighbor algorithm \cite{cover1967nearest} is a
well-studied technique for classification and approximation. Nearest
neighbor inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi
cell \cite{dirichlet1850reduction}) consisting of all points closer to
$x^{(i)}$ than to any other point $x^{(j)}$. The Voronoi model smooths
the nearest neighbor approximation by utilizing the Voronoi cells to
define support via a generic basis function $v: \mathbb{R}^d
\rightarrow \mathbb{R}_+$ given by
 $$ v^{x^{(i)}}(y) = \left(1 - \frac{\bigl\|y - x^{(i)}\bigr\|_2}
 {2 \ h \bigl(y - x^{(i)} \mid x^{(i)}\bigr)} \right)_+, $$
where $h\bigl(w \mid x^{(i)}\bigr)$ is the distance between $x^{(i)}$
and the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction
$w$. $v^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $v^{x^{(i)}}$
has local support, giving the interpolated value
 $$ f(y) = \frac{\sum\limits_{i=1}^n v^{x^{(i)}}(y)
  f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n v^{x^{(i)}}(y)}, $$
where $0 \leq v^{x^{(i)}}(y) \leq 1$. The computational complexity of
interpolation via this Voronoi model is $\mathcal{O}(n^2 d)$. All of
the approximations are an interpolant involving a convex combination
of known function values $f\bigl(x^{(i)}\bigr)$.

@@

\changed{
\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth,]{example_distribution_prediction.pdf}
  \caption{In this HPC I/O example, the general methodology for
    predicting a CDF and evaluating error can be seen\color{red},
    where M means $\times 10^6$\color{black}. The Delaunay method
    chose three source distributions (dotted lines) and assigned
    weights \color{red}\{.1, .3, .6\} \color{black} (top to bottom at
    \color{red}middle\color{black}). The weighted sum of the three
    known CDFs produces the predicted CDF (dashed line). The KS
    Statistic (\color{red}vertical line\color{black}) computed between
    the true CDF (solid line) and predicted CDF (dashed line) is 0.2
    for this example. For this example the KS test null hypothesis is
    rejected at $p$-value 0.01, however it is not rejected at
    $p$-value 0.001.
  \vspace{-.1cm}}
  \label{fig:prediction-example}
\end{figure}
}{}

@@

A hurdle when modeling function-valued outputs such as cumulative
distribution functions (CDFs) or probability density functions (PDFs)
is that certain properties must be maintained. It is necessary that a
PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x)
\geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. \changed{Instead}{However}, for a CDF
$F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are
$F(x) \in [0,1]$ and $F(x)$ is absolutely continuous and
nondecreasing. This work utilizes the fact that a convex combination
of CDFs (or PDFs) results in a valid CDF (or PDF). Given $G(x) =
\sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$
is a valid CDF, $G$ must also be a valid CDF. A demonstration of how
this is applied can be seen in Figure \ref{fig:prediction-example}. 
\changed{In this example the KS test null hypothesis is rejected at $p$-value
0.01, however it is not rejected at $p$-value 0.001.}{}

@@

The two-sample KS test is a useful nonparametric test for comparing
two \changed{empirical}{} CDFs while only assuming stationarity,
finite mean, and finite variance. The null hypothesis (that two
\changed{empirical}{} CDFs come from the same underlying distribution)
is rejected at level $p \in [0,1]$ when
 $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)}
\sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$ with distribution sample
sizes $n_1,n_2 \in \mathcal{N}$. For all applications of the KS test
presented in this work $n_1 = n_2$. An example of \changed{the process
  of generating predicted distributions from known distributions and
  the subsequent}{the round-trip prediction methodology from known and
  predicted distributions to the} calculation of error can be seen in
Figure \ref{fig:prediction-example}. A brief listing of relevant
statistical terms used throughout this work is provided in the
Appendix (Section \ref{sec:appendix}).

@@

\begin{proofdot}
  \def\g{\gamma_g}
  Consider $g(t)=f(x(t)) - \hat f(x(t))$ along
  the line segment $x(t)=(1-t)x_0 + tx_j$, $0\le t\le1$, from $x_0$ to
  $x_j$.  Observe that $g(0)=f(x_0) - \hat f(x_0) = 0$, $g(1)=f(x_j) -
  \hat f(x_j) = 0$, and $g'(t)$ is $\g$-Lipschitz continuous with $\g
  = \gamma \|x_j-x_0\|_2^2$. \changed{The following is visualized in 
  Figure \ref{fig:lemma3}.}{}

>>>>

%% %% Include the response to reviewers at the beginning of the document.
%% \include{tchlux-NA-response-2_2020-04}

@@

Regression and interpolation have a considerable theoretical base in
one dimension \cite{cheney2009course}. Common techniques include
Lagrangian interpolation, piecewise linear interpolation, and more
generally spline interpolation via B-splines
\cite{de1978practical}. Tensor products of B-splines
\cite{unther1996interpolating} or other basis functions have an
unfortunate exponential scaling with increasing dimension. Exponential
scaling prohibits tensor products from being reasonably applied beyond
three-dimensional data. In order to address this dimensional scaling
challenge, C. de Boor and others proposed box splines
\cite{de2013box}, of which one of the approximation techniques in this
work is composed \cite{lux2018novel}.

The theoretical foundation of low dimensional interpolation allows the
construction of strong error bounds that are lacking in high
dimensional problems. A large body of literature exists studying the
construction and application of sparse grids
\cite{bungartz2004sparse,chkifa2014high}, which attempt to address the
problem of dimensional scaling. The sparse grid point sets combined
with specific approximation functions allow for the construction of
interpolation error bounds as well as bounds on total error for
regressors by considering projections onto orthogonal polynomials
\cite{nobile2016convergence}. However, sparse grids are not studied
here because their size still grows quite fast with dimension
(although less than exponentially) making them not readily tractable
for more than a few tens of dimensions and are primarily used for
problems that allow data points to be chosen. This work extends some
known results regarding the secant method \cite{dennis1996numerical}
to construct a uniform error bound for interpolants over arbitrary
scattered data in any dimension. These error bounds evidence that
interpolants are useful, considering similar uniform bounds cannot be
constructed for regressors in general
\cite{migliorati2013approximation}. The maximum complexity of an
interpolant is bounded by the amount of data available, while the
maximum complexity of a regressor is bounded by both the amount of
data and the chosen parametric form. For this reason, generic uniform
bounds are largely unobtainable for regression techniques on arbitrary
approximation problems, even when the approximation domain is bounded.

@@

The most challenging problem when scaling in dimension is that the
number of possible interactions between dimensions grows
exponentially. Quantifying all possible interactions becomes
intractable, and hence problems beyond ten to twenty dimensions
usually rely on linear models. That is not to say nonlinear models are
absent, but nonlinearities are often either preconceived or model
interactions between small subsets of dimensions at most. Even
globally nonlinear approximations such as neural networks are
constructed from compositions of summed low-interaction functions
\cite{clevert2015fast}.

@@

\subsection{Multivariate Adaptive Regression Splines}
\label{sec:mars}
This approximation was introduced in \cite{friedman1991multivariate}
and subsequently improved to its current version in
\cite{stanford1993fast}, called fast multivariate adaptive regression
splines (Fast MARS). In Fast MARS, a least squares fit model is
iteratively built by beginning with a single constant valued function
and adding two new basis functions at each iteration of the form
\begin{align*}
  B_{2j-1}(x) &= B_l(x) \bigl(x_i-x^{(p)}_i\bigr)_+, \\
  B_{2j}(x) &= B_k(x) \bigl(x_i-x^{(p)}_i\bigr)_- ,
\end{align*}
where $j \leq m$ is the iteration number, $m$ is the maximum
  number of underlying basis functions, $1 \le p \le n$, and
$B_l(x)$, $B_k(x)$ are basis functions from the previous iteration,
 $$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$

$w_- = (-w)_+$, and $i$, $p$, $l$, and $k$ are chosen to produce the
best fit to the data with the addition of $B_{2j-1}$ and $B_{2j}$ to
the existing basis, among all possible choices of $i$, $p$, $l$, and
$k$. After iteratively constructing a model, MARS then iteratively
removes basis functions that do not contribute to goodness of fit. In
effect, MARS creates a locally component-wise tensor product
approximation of the data. The overall computational complexity of
Fast MARS is $\mathcal{O}(n d m^3)$. See the work of Friedman
\cite{stanford1993fast} for a more thorough description of this
approximation and its construction. This paper uses an
implementation of Fast MARS \cite{rudy2017pyearth} with $m = 200$
throughout all experiments.

@@

\subsection{Multilayer Perceptron Regressor}
\label{sec:mlp}
The neural network is a well studied and widely used method for both
regression and classification tasks
\cite{hornik1989multilayer,rumelhart1988learning}. When using the
rectified linear unit (ReLU) activation function
\cite{dahl2013improving} and fitting the model with stochastic
gradient descent (SGD) or BFGS minimization techniques
\cite{goh2017why,moller1993scaled,robbins1951stochastic}, the model
built by a multilayer perceptron uses layers $l : \mathbb{R}^{i}
\rightarrow \mathbb{R}^{j}$ defined by
$$ l(u) = \bigl( u^t W_l + c_l \bigr)_+ ,$$
where $c_l \in \mathbb{R}^j$ and $W_l$ is the $i \times j$ weight
matrix for layer $l$. In this form, the multilayer perceptron (MLP)
produces a piecewise linear model of the input data. The computational
complexity of training a multilayer perceptron is $\mathcal{O}(n d
m)$, where $m$ is determined by the sizes of the layers of the network
and the stopping criterion of the minimization used for finding
weights. This paper uses an MLP built from Keras and Tensorflow to
perform regression \cite{tensorflow2015-whitepaper,chollet2015keras}
with ten hidden layers each having thirty two nodes (a total of
approximately ten thousand parameters), ReLU activation, and one
hundred thousand steps of SGD for training. It should be noted that
the choice of neural network architecture affects approximation
performance, but no architecture tuning is performed here.

@@

The Delaunay triangulation is a well-studied geometric technique for
producing an interpolant \cite{lee1980two}. The Delaunay triangulation
of a set of points into simplices is such that there
are no points inside the sphere defined by the
vertices of each simplex, visualized in Figure \ref{fig:delaunay}.
For a $d$-simplex S with vertices $x^{(0)}$,
$x^{(1)}$, $\ldots$, $x^{(d)}$, and function values
$f\bigl(x^{(i)}\bigr)$, $i=0$, $\ldots$, $d$, $y \in S$ is a unique
convex combination of the vertices,
 $$ y = \sum_{i=0}^{d} w_i x^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad w_i \geq 0, \quad i=0,\ldots,d, $$
and the Delaunay interpolant $\hat f(y)$ at $y$ is given by
 $$ \hat f(y) = \sum_{i=0}^{d} w_i f\bigl(x^{(i)}\bigr). $$ The
computational complexity of Delaunay interpolation (for the
implementation used \cite{chang2018polynomial}) is $\mathcal{O}(k n
d^3)$. Given pathological data the entire triangulation could be
computed with $k = n^{\lceil d / 2 \rceil}$, but the average case yields
$k = d \log d$ and is capable of scaling reasonably to $d \leq 50$. In
the present application, a Delaunay simplex $S$ containing $y$ is
found, then the $d+1$ vertices (points in $X$) of $S$ are used to
assign weights to each vertex and produce the predicted function value
for point $y$.

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth,]{example_delaunay.pdf}
  \includegraphics[width=0.4\textwidth,]{example_not_delaunay.pdf}
  \caption{ On the left above is a depiction of a Delaunay
    triangulation over four points, notice that the circumball (shaded
    circle) for the left simplex does not contain the fourth point.
    On the right above, a non-Delaunay mesh is depicted. Notice that
    the circumball for the top simplex (shaded circle, clipped at
    bottom edge of the visual) contains the fourth point which
    violates the Delaunay condition for a simplex.
  \vspace{-.1cm}}
  \label{fig:delaunay}
\end{figure}

@@

The modified Shepard method used here (ShepMod) generates a continuous
approximation based on Euclidean distance and resembles a nearest
neighbor interpolant \cite{cover1967nearest}. This model is a type of
\textit{metric interpolation}, also called a Shepard method
\cite{gordon1978shepard,shepard1968two}. The interpolant has the form
$$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)} {\sum\limits_{k=1}^{n}W_k(x)} ,$$ 
where $p\bigl(x^{(k)}\bigr) = f\bigl(x^{(k)}\bigr)$, $W_k(x) =
\bigl(\bigl(r_k - \bigl\|x - x^{(k)}\bigr\|_2\bigr)_+ \big/ \bigl( r_k
\bigl\|x - x^{(k)}\bigr\|_2 \bigr)\bigr)^2$ for $x \neq x^{(k)}$, and
$r_k \in \mathbb{R}$ is the smallest radius such that at least $d+1$
other points $x^{(j)}$, $j \not = k$, are inside the closed Euclidean
ball of radius $r_k$ about $x^{(k)}$. The interpolant $p(x)$ is
continuous because the singularities at the $x^{(k)}$ are
removable. The weight function is visualized in Figure \ref{fig:shepard}.
The computational complexity of ShepMod is $\mathcal{O}(n^2d)$. This
paper uses a Fortran 95 implementation of ShepMod derived from
SHEPPACK \cite{thacker2010algorithm}.

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth,]{example_shepard_radius.pdf}
  \hspace{.2cm}
  \includegraphics[width=0.4\textwidth,]{example_shepard_weight.pdf}
  \caption{On the left above is a depiction of the radius
    of influence for three chosen points of a collection in two
    dimensions using the modified Shepard criteria. On the right a
    third axis shows the relative weight for the center most
    interpolation point $x^{(i)}$ with the solid line representing its
    radius of influence, where $W_i(x)$ is $0$ for $\|x-x^{(i)}\|_2
    \geq r_i$ and $W_i(x) / \sum_{k=1}^n W_k(x) \to 1$ as $x \to
    x^{(i)}$.
  \vspace{-.1cm}}
  \label{fig:shepard}
\end{figure}

\subsection{Linear Shepard}
The linear Shepard method (LSHEP) is a blending function using local
linear interpolants, a special case of the general Shepard algorithm
\cite{thacker2010algorithm}. The interpolant has the form
 $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)P_k(x)}
 {\sum\limits_{k=1}^{n}W_k(x)} ,$$
where $W_k(x)$ is the same as for the modified Shepard method and
$P_k(x)$ is a local linear approximation to the data satisfying
$P_k\bigl(x^{(k)}\bigr) = f\bigl(x^{(x)}\bigr)$. The computational
complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This paper uses the
Fortran 95 implementation of LSHEP in SHEPPACK
\cite{thacker2010algorithm}.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth,]{VM_Construction.pdf}
  \caption{Above is a depiction of the Voronoi cell
    boundaries (dashed lines) about a set of interpolation points
    (dots) in two dimensions. In this example, the Voronoi model basis
    function about the center most point has nonzero weight in the
    shaded region and transitions from a value of one at the point to
    zero at the boundary of the twice expanded Voronoi cell (solid
    line).
  \vspace{-.1cm}}
  \label{fig:voronoi-basis}
\end{figure}

The nearest neighbor algorithm \cite{cover1967nearest} is a
well-studied technique for classification and approximation. Nearest
neighbor inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi
cell \cite{dirichlet1850reduction}) consisting of all points closer to
$x^{(i)}$ than to any other point $x^{(j)}$. The Voronoi model smooths
the nearest neighbor approximation by utilizing the Voronoi cells
(visualized in Figure \ref{fig:voronoi-basis}) to
define support via a generic basis function $v: \mathbb{R}^d
\rightarrow \mathbb{R}_+$ given by
 $$ v^{x^{(i)}}(y) = \left(1 - \frac{\bigl\|y - x^{(i)}\bigr\|_2}
 {2 \ h \bigl(y - x^{(i)} \mid x^{(i)}\bigr)} \right)_+, $$
where $h\bigl(w \mid x^{(i)}\bigr)$ is the distance between $x^{(i)}$
and the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction
$w$. $v^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $v^{x^{(i)}}$
has local support, giving the interpolated value
 $$ f(y) = \frac{\sum\limits_{i=1}^n v^{x^{(i)}}(y)
  f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n v^{x^{(i)}}(y)}, $$
where $0 \leq v^{x^{(i)}}(y) \leq 1$. The computational complexity of
interpolation via this Voronoi model is $\mathcal{O}(n^2 d)$. All of
the approximations are an interpolant involving a convex combination
of known function values $f\bigl(x^{(i)}\bigr)$.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth,]{example_distribution_prediction.pdf}
  \caption{In this HPC I/O example, the general methodology for
    predicting a CDF and evaluating error can be seen, where M means
    $\times 10^6$. The Delaunay method chose three source
    distributions (dotted lines) and assigned weights \{.1, .3, .6\}
    (top to bottom at middle). The weighted sum of the three known
    CDFs produces the predicted CDF (dashed line). The KS Statistic
    (vertical line) computed between the true CDF (solid line) and
    predicted CDF (dashed line) is 0.2 for this example. For this
    example the KS test null hypothesis is rejected at $p$-value
    threshold 0.01, however it is not rejected at $p$-value threshold
    0.001.
  \vspace{-.1cm}}
  \label{fig:prediction-example}
\end{figure}

@@

A hurdle when modeling function-valued outputs such as cumulative
distribution functions (CDFs) or probability density functions (PDFs)
is that certain properties must be maintained. It is necessary that a
PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x)
\geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. Instead, for a CDF
$F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are
$F(x) \in [0,1]$ and $F(x)$ is absolutely continuous and
nondecreasing. This work utilizes the fact that a convex combination
of CDFs (or PDFs) results in a valid CDF (or PDF). Given $G(x) =
\sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$
is a valid CDF, $G$ must also be a valid CDF. A demonstration of how
this is applied can be seen in Figure \ref{fig:prediction-example}.

@@

The two-sample KS test is a useful nonparametric test for comparing
two empirical CDFs while only assuming stationarity,
finite mean, and finite variance. The null hypothesis (that two
empirical CDFs come from the same underlying distribution)
is rejected at level $p \in [0,1]$ when
$$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
with distribution sample sizes $n_1,n_2 \in \mathbb{N}$
(\textbf{not} the number of approximation points). For all
applications of the KS test presented in this work $n_1 = n_2$. An
example of the process of generating predicted distributions from
known distributions and the subsequent calculation of error can be
seen in Figure \ref{fig:prediction-example}. In this example the KS
test null hypothesis is rejected at $p$-value threshold 0.01, however it is not
rejected at $p$-value threshold 0.001. A brief listing of relevant statistical
terms used throughout this work is provided in the Appendix (Section
\ref{sec:appendix}).

@@

\begin{proofdot}
  \def\g{\gamma_g}
  Consider $g(t)=f(x(t)) - \hat f(x(t))$ along
  the line segment $x(t)=(1-t)x_0 + tx_j$, $0\le t\le1$, from $x_0$ to
  $x_j$.  Observe that $g(0)=f(x_0) - \hat f(x_0) = 0$, $g(1)=f(x_j) -
  \hat f(x_j) = 0$, and $g'(t)$ is $\g$-Lipschitz continuous with $\g
  = \gamma \|x_j-x_0\|_2^2$. The following is visualized in 
  Figure \ref{fig:lemma3}.

  Suppose $g'(0) > \g/2$. Then $|g'(0)-g'(t)| \le \g t \implies g'(0)
  - \g t \le g'(t)$, and the line $w=g'(0)-\g t$ intersects the
  $t$-axis at $\tilde t=g'(0)/\g > 1/2$. $\tilde t<1$ necessarily,
  since by Rolle's Theorem there exists $0<z<1$ such that $g'(z)=0$
  and so $g'(0) - \g z \le g'(z) = 0$. Now by integrating
  $$\frac{g'(0)^2}{2\g} = \frac{g'(0)\tilde t}{2} = \int_0^{\tilde t}
  (g'(0)-\g t)dt \le \int_0^{\tilde t} g'(t)dt = g(\tilde t),$$ and
  using $g'(0) > \g / 2$
  \begin{align*}
    \frac{-g'(0)^2}{2\g} < g'(0) - \frac{g'(0)^2}{2\g} - \g/2 =
    \frac{(1-\tilde t)(g'(0)-\g)}{2} &= \int_{\tilde t}^1 (g'(0)-\g
    t)dt \\ &\le \int_{\tilde t}^1 g'(t)dt = -g(\tilde t),
  \end{align*}
  \noindent a contradiction for the value of $g(\tilde t)$. A similar
  contradiction arises for $g'(0)<-\g/2$. Therefore $|g'(0)| \le
  \g/2$. In terms of $f$,
  $$\bigl|\langle \nabla f(x_0)-c,x_j-x_0\rangle\bigr| = |g'(0)| \le
  \gamma \|x_j-x_0\|_2^2/2 \le \gamma k^2 / 2,$$ which holds for all
  $1\le j\le d$.  Finally, using {\it Lemma \ref{lemma:2}},
  $$\|\nabla f(x_0)-c\|_2^2 \le \frac{d}{\sigma_d^2}(\gamma k^2/2)^2
  \implies \|\nabla f(x_0)-c\|_2 \le \frac{\sqrt{d} \, \gamma
    k^2}{2\sigma_d}.$$ \qed
\end{proofdot}

\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth,]{example_lemma_3.pdf}
  \def\g{\gamma_g}
  \caption{ Three different scenarios visualizing {\it Lemma
      \ref{lemma:3}}, where $g(t)$ is the difference between a
    piecewise linear interpolant and the approximated function along a
    normalized line segment between interpolation points, $g'(t)$ is
    $\g$-Lipschitz continuous, and $w$ and $\tilde t$ are defined in
    the proof. Leftmost is a randomly chosen permissible shape of $g$
    and $g'$. The middle is the only possible shape of $g$ and $g'$
    given $g'(0) = \g / 2$, establishing the case of equality in the
    lemma. Rightmost is the resulting contradiction when $g'(0) > \g /
    2$, notice it is impossible to ensure $g'(t)$ is $\g$-Lipschitz
    continuous and satisfy $g(1) = 0$ (highlighted with red circle on
    the right).
  \vspace{-.1cm}}
  \label{fig:lemma3}
\end{figure}
